---
title: "<span style='font-size: 16px; color: grey'> crisprQTL proof-of-concept experiment in T cells </span> <br> <span style='font-size: 36px'> Visualisation of perturbation effects </span>"
author: "Ximena Ibarra-Soria"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  html_document:
    theme: paper
    keep_md: true
    fig_width: 5
    fig_height: 5
    fig_caption: yes
    df_print: paged
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(scran)
library(scales)
library(ggplot2)
library(ggpubr)

dir <- paste0(getwd(), "/")

th <- theme_bw() + theme(
  axis.text.x = element_text(size=10), 
  axis.title.x = element_text(size=12), 
  axis.text.y = element_text(size=12), 
  axis.title.y = element_text(size=12),
  panel.grid.major = element_blank(), 
  panel.grid.minor = element_blank(), 
  axis.line = element_line(colour = "black"), 
  panel.border = element_blank(), 
  plot.title = element_text(face="bold", hjust = 0.5))

## FUNCTIONS

plot_gene_sc <- function(sce=NULL, cells.neg=NULL, hit=NULL, thr=0.05){
  ## get relevant data
  gene <- hit$gene_id
  target <- hit$target
  guides <- unlist(strsplit(hit$gRNAs, ",", fixed=TRUE))
  pvals <- as.numeric(unlist(strsplit(hit$p_values, ",", fixed=TRUE)))
  names(pvals) <- guides
  fcs <- as.numeric(unlist(strsplit(hit$logFCs, ",", fixed=TRUE)))
  names(fcs) <- guides
  
  ## get cells with gRNAs in `guides`
  cells.guides <- sapply(guides, function(x) names(which(counts(altExp(sce, 'gRNA_calls'))[x,])), simplify = FALSE)
  cells.pos <- rep(guides, times=unlist(lapply(cells.guides, length)))
  names(cells.pos) <- unname(do.call(c, cells.guides))
  
  ## retrieve expression of `gene`
  df <- data.frame(cell = union(names(cells.pos), cells.neg))
  df$expr <- logcounts(sce[,df$cell])[gene,]
  # label perturbed vs background
  df$status = factor(ifelse(df$cell %in% names(cells.pos), "perturbed", "background"), levels=c("perturbed", "background"))
  # annotate each gRNA
  df <- cbind(df, sapply(unique(cells.pos), function(x) df$cell %in% names(cells.pos[cells.pos==x])))
  # remove cells with several on-target gRNAs (to keep plots consistent and avoid having some with more violins than others)
  if(ncol(df)>4){
    df$guide <- ifelse(rowSums(df[,-c(1:3)])>1 , "multiple",
                       ifelse(rowSums(df[,-c(1:3)])==0, "background", "guide"))
    df <- df[df$guide != "multiple",]
    for(g in unique(cells.pos)){
      df[df[,g]==TRUE & df$guide == "guide",]$guide <- g
    }
    df$guide <- factor(df$guide, levels=c(unique(cells.pos), "background"))
  }else{
    df$guide <- ifelse(df[,4]==TRUE, colnames(df)[4], "background")
    df$guide <- factor(df$guide, levels=c(colnames(df)[4], "background"))
  }
  
  ## colour based on significance per-guide
  df$colour <- ifelse(df$status == "background", "background", "perturbed")
  
  ## plot
  p <- ggplot(df, aes(guide, expr, colour = colour)) +
    geom_violin(lwd=1) +
    geom_jitter(stat="identity",
                position=position_jitter(0.1), 
                size=1, alpha=0.2) +
    scale_color_manual(values = c(perturbed="lightcoral", background="grey60")) +
    geom_boxplot(outlier.shape = NA, colour="grey20", width=0.1, fill=NA) +
    geom_hline(yintercept = median(df[df$status=="background",]$expr), colour="grey20", lty=2) +
    xlab("") + 
    ylab(expression('log'[2]*' normalised counts')) +
    ggtitle(label = hit$pair_tgt, 
            subtitle = paste("FDR:", formatC(hit$FDR, format = "e", digits = 3), 
                             "\t\ttier:", toupper(hit$tier), "\t\tdistance:", round(hit$distance/1e3, 1), "kb\n")) +
    th + theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5),
               legend.position = "none")

  ## annotate
  p <- p + annotate("text", 
                    x = 1:length(levels(df$guide)),
                    y = max(df$expr)+((max(df$expr)-min(df$expr))*0.3), 
                    label = paste0(
                      c(formatC(pvals[setdiff(levels(df$guide),"background")], format = "e", digits = 2), ""), "\n",
                      c(round(fcs[setdiff(levels(df$guide),"background")], 2), ""), "\n",
                      paste0("(", as.numeric(round(prop.table(table(df$guide, df$expr>0),1)*100,1)[,'TRUE']), ")"), "\n",
                      as.numeric(c(table(cells.pos)[setdiff(levels(df$guide),"background")], 5e3))),
                    size = 3, 
                    colour = c(ifelse(pvals[setdiff(levels(df$guide),"background")] < thr, "lightcoral", "grey40"), "grey40")) +
        coord_cartesian(clip = "off")

  return(p)
}

```

We have determined the set of genes that change expression upon perturbation of each target locus. Here, we plot the expression changes for all significant results.

```{r data}
## data
sce <- readRDS(paste0(dir, "results/03_sce_POC_Tcells.goodQual.calls.NORM.Rds"))

## gRNA annotation
gRNA_ann <- read.table(paste0(dir, "data/gRNA_library.tsv"), sep="\t", header = TRUE)

## DEGs
degs <- read.table(paste0(dir, "results/06_DEgenes.tsv"), header = TRUE)
# all DE testing results
res <- read.table(paste0(dir, "results/06_DEresults_targetLevel.tsv"), header = TRUE)
res$tier <- degs[match(res$pair_tgt, degs$pair_tgt),'tier']
res$distance <- degs[match(res$pair_tgt, degs$pair_tgt),'distance']

# define NT cells for plotting
nt_guides <- gRNA_ann[gRNA_ann$class=="NT",]$ID
cells <- colnames(sce[,sce$call == "unique"])
cells.neg <- apply(counts(altExp(sce, 'gRNA_calls'))[nt_guides, cells], 1, function(x) names(which(x)))
cells.neg <- unlist(cells.neg)
# use 5K
set.seed(749)
cells.neg <- sample(cells.neg, size=5e3)
```


### True positive effects {.tabset}

First, we plot the expression of the genes that are expected to be downregulated, from positive control perturbations.

#### TSS

```{r TP_tss, message=FALSE, warning=FALSE, fig.width=10, fig.height=68}
selected <- setdiff(unique(gRNA_ann[gRNA_ann$class=="TSS",]$target), c("IL12A_TSS", "TBX1_TSS")) # IL12A and TBX1 are not expressed

## plot every target
plots <- list()
for(target in selected){
  # expected gene
  gene <- unique(gRNA_ann[gRNA_ann$target==target,]$expected_DE_gene)

  # DE results
  hit <- res[res$target==target & res$gene_id==gene,]

  # plot
  plots[[target]] <- plot_gene_sc(sce = sce, cells.neg = cells.neg, hit = hit, thr=0.05)
}
ggarrange(plotlist = plots, ncol=2, nrow=17, align="hv")
```


#### LCR

```{r TP_LCR, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
selected <- unique(gRNA_ann[gRNA_ann$class=="LCR",]$target)

## plot every target
plots <- list()
for(target in selected){
  # expected gene
  gene <- unique(gRNA_ann[gRNA_ann$target==target,]$expected_DE_gene)
  
  # DE results
  hit <- res[res$target==target & res$gene_id==gene,]
    
  # plot
  plots[[target]] <- plot_gene_sc(sce = sce, cells.neg = cells.neg, hit = hit, thr=0.05)
}
ggarrange(plotlist = plots, ncol=2, nrow=2, align="hv")
```


#### ENH

```{r TP_enh, message=FALSE, warning=FALSE, fig.width=10, fig.height=52}
selected <- setdiff(unique(gRNA_ann[gRNA_ann$class=="ENH",]$target), c("USP6NL_ENH", "TUBB2A_ENH")) # USP6NL and TUBB2A are not expressed

## plot every target
plots <- list()
for(target in selected){
  # expected gene
  gene <- unique(gRNA_ann[gRNA_ann$target==target,]$expected_DE_gene)

  # DE results
  hit <- res[res$target==target & res$gene_id==gene,]
    
  # plot
  plots[[target]] <- plot_gene_sc(sce = sce, cells.neg = cells.neg, hit = hit, thr=0.05)
}
ggarrange(plotlist = plots, ncol=2, nrow=13, align="hv")
```



### Additional hits {.tabset}

Below are any other **high- and medium-confidence hits**, that are not the expected gene for intergenic perturbations.

#### ENH

```{r other_enh, message=FALSE, warning=FALSE, fig.width=10, fig.height=84}
## plot every target
selected <- degs[degs$class=="ENH" & !degs$expected & degs$tier!="low",]

plots <- list()
for(i in 1:nrow(selected)){
  hit <- selected[i,]

  plots[[i]] <- plot_gene_sc(sce = sce, cells.neg = cells.neg, hit = hit, thr=0.05)
}
ggarrange(plotlist = plots, ncol=2, nrow=21, align="hv")
```


#### INTRON

```{r other_intron, message=FALSE, warning=FALSE, fig.width=10, fig.height=48}
## plot every target
selected <- degs[degs$class=="INTRON" & degs$tier!="low",]

plots <- list()
for(i in 1:nrow(selected)){
  hit <- selected[i,]

  plots[[i]] <- plot_gene_sc(sce = sce, cells.neg = cells.neg, hit = hit, thr=0.05)
}
ggarrange(plotlist = plots, ncol=2, nrow=12, align="hv")
```

#### INTERGENIC

```{r other_intergenic, message=FALSE, warning=FALSE, fig.width=10, fig.height=24}
## plot every target
selected <- degs[degs$class=="INTERGENIC" & degs$tier!="low",]

plots <- list()
for(i in 1:nrow(selected)){
  hit <- selected[i,]

  plots[[i]] <- plot_gene_sc(sce = sce, cells.neg = cells.neg, hit = hit, thr=0.05)
}
ggarrange(plotlist = plots, ncol=2, nrow=6, align="hv")
```

### Overlapping gene for intronic perturbations

For intronic elements, we check directly the expression of the overlapping gene. *LCA5L* is not detected in at least 5% of cells and was not tested.

```{r intron_gene, message=FALSE, warning=FALSE, fig.width=10, fig.height=20}
## plot every target
df <- data.frame(target = unique(gRNA_ann[gRNA_ann$class=="INTRON",]$target),
                 gene = unlist(lapply(strsplit(unique(gRNA_ann[gRNA_ann$class=="INTRON",]$target), "_"),'[[',1)))
df <- df[-which(df$target == "LCA5L_INTRON"),]

plots <- list()
for(i in 1:nrow(df)){
  target <- df[i,1]
  gene <- df[i,2]
  hit <- res[res$target == target & res$gene_id == gene,]
  
  plots[[i]] <- plot_gene_sc(sce = sce, cells.neg = cells.neg, hit = hit, thr=0.05)
}
ggarrange(plotlist = plots, ncol=2, nrow=5, align="hv")
```


### Nearby genes for intergenic perturbations

We also plot the expression of the four closest genes (detected in at least 5% of the cells) for `INTERGENIC` targets.

```{r intergenic_nearby, message=FALSE, warning=FALSE, fig.width=10, fig.height=24}
## neighbourhood
gene_distance_target <- read.table(paste0(dir, "results/06_gene_target_distance.tsv"), header=TRUE)

## plot every target
df <- gene_distance_target[grep("INTERGENIC", gene_distance_target$target),]

plots <- list()
for(target in unique(gRNA_ann[gRNA_ann$class=="INTERGENIC",]$target)){
  selected <- df[which(df$target==target & df$expressed),][1:4,]
  for(i in 1:nrow(selected)){
    hit <- res[res$target == target & res$gene_id == selected[i,'external_gene_name'],]
    plots[[paste0(target,i)]] <- plot_gene_sc(sce = sce, cells.neg = cells.neg, hit = hit, thr=0.05)
  }
}
ggarrange(plotlist = plots, ncol=2, nrow=6, align="hv")
```


```{r info}
sessionInfo()
```

